---
title: "Participatory Systems Dynamics Facilitation Guide: Dynamics"
author: "Swap Mushiana, M.S."
date: "May 13, 2017"
output:
  html_document: default
  ioslides_presentation: default
  mode: self-contained
  pdf_document: default
  slidy_presentation: default
  word_document: default
---

### Participatory System Dynamics Resources in R Markdown 

This is an R Markdown presentation for the "Ratio Exercise" learning loop exercise from the participatory learning session of the Team PSD Facilitation Guide. 

Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. 

We use R and R Markdown so that our materials are open source/free, transparent for collaboration, reproducible by others and can be used repeatedly over time for local sustainability or for adaptation to another local health care system.

### Draft: Dynamics One-Pager
##### Last edited: 5.13.17
```{r}
```
#### Key Idea(s) -  Lesson Hexagon 
* An enhanced understanding of dynamics will empower stakeholders to make informed change-decisions within their system. 
* Increase stakeholder ability to use feedback thinking to link service loads to mental health resources in VHA.
* How ratio effects positive and negative feedback loops of a system over time. 
```{r}
```
#### Definitions 
* **System Behavior – Oscillation** 
    + The “dated” information to make decisions or control a target is likely to miss or overshoot its goal
    + Dated: control action is not based on *current* state of system, but rather a *previous state or value.* 

* **Static Observation** -- DEFINITION NEEDED

* **Dynamic Observation over time- Key to understanding drivers**
    + *Endogenous variables* that are influenced by other variables in a system 
    + *Exogenous variables* in the model that are strictly causes of other variables and not influenced by other variables in a model 
    + *Dynamic Complexity* refers to the Number of dynamic behavior patterns that a system can produce (e.g., oscillations, overshoot and collapse, etc.) 
    
 * **Non-linear System Behavior - Exponential Growth**
    + Growth can occur suddenly after a long period of little or no change.Example: folding a peice of paper 36 plus times. 
   
 * **Non-linear System Behavior - Steady State**
    + *Equilibrium point* that attracts or repels the state of the system regardless of the initial state.
    + Commonly held belief during steady state, *“The clinic has reached its capacity for improvement.”*


```{r}
```
#### Dialogue - Hexagon 
* What variables within a system 
* How do those variables behave or change over time?
* What tool should we use to explain the dynamics that we believe are important to the system? (A model  perhaps???)

```{r}

```
#### Learning Loop (Practice/Exercise)
* Mention Alexandra’s “zoom-in” and “zoom-out” skill to look at how static observation can miss important dynamic observations such as oscillation and nonlinear system behaviors.
* Touch on Myra’s skill and build off the exercise she introduces with casual loops. 
* **Ratio Exercise**
    + Collaborate with stakeholders to map ratios (or other language used) with + or – feedback loops. 
    + Prompt stakeholders to tell “feedback stories” to a “what if” scenarios proposed by facilitator. 
```{r}
```
#### Lesson Overview - Hexagon Molecules
* Stand-alone sentence, one sentence summary: NEED ONE SENTENCE HERE
* Skills: Increase stakeholders’ capacity for feedback thinking for the relationships amongst service loads (Veteran’s need for EBPs) and mental health resources (scheduling, # of providers, types of services offered, etc.)
* PSD uses specific tool, a model, to depict the dynamics of a system into a model to inform clinical level decisions . 

```{r}
```
#### Lesson Complete - Hexagon Badge
```{r}
```
#### Resources 
1. Scriptapedia 4.0.6 (2013). Original authors are David F. Andersen and George P. Richardson.
2. http://www.systemdynamics.org/DL-IntroSysDyn/exgr.htm 

